Tidy data structure to support exploration and modeling of temporal-context data

### 2018-04-28 main.Rmd
added hts introduction

---
title: "Tidy data structure to support exploration and modeling of temporal-context data"
author:
  - familyname: Wang
    othernames: Earo
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : earo.wang@monash.edu
    correspondingauthor: true
  - familyname: Cook
    othernames: Dianne
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : dicook@monash.edu
  - familyname: Hyndman
    othernames: Rob J
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : rob.hyndman@monash.edu
abstract: 'Temporal-context data is often rich with information and time formats, for example multiple observational units, different time lengths, heterogeneous data types, nested and crossed factors and etc. This work presents a cohesive and conceptual framework for organizing and manipulating temporal data, which in turn flows into visualization and modelling routines. Tidy data principles are applied and extended to temporal data: (1) mapping the semantics of a dataset into its physical layout, (2) an explicitly declared index variable representing time, (3) a "key" comprised of single or multiple variables to uniquely identify units over time, using a syntatical and user-oriented approach in which it imposes nested or crossed structures on the data. This tidy data representation most naturally supports thinking of operations on the data as building blocks, forming part of a "data pipeline" in time-based context. A sound pipeline practice facilitates a transparent and human-readable workflow for analyzing temporal data. Applications are included to illustrate tidy temporal data structure, data pipeline ideas and usage. The architecture of tidy temporal data has been implemented in the R package **tsibble**.'
keywords: "temporal context, time series, data structure, R"
wpnumber: no/yr
jelcodes: C10,C14,C22
blind: false
cover: true
toc: false
bibliography: references.bib
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: no
    number_sections: yes
    citation_package: biblatex
---

@@

However, such richness in temporal data cannot be easily accommodated by currently available data structures in any statistical computing package. In the literature, time series and panel (longitudinal) data are common terms referred to temporal-context data, depending on the research fields. Researchers who are concerned with modelling large $T$ and small $N$ would name as "time series" (serial correlation); those who are interested in modelling small $T$ and large $N$ as "panel data" (asymptotic). The data format is two-dimensional array, but different modelling focuses lend the data input to column types. A matrix is used to represent multivariate time series where each row represents observations measured at a time point and each column represents a series. This matrix representation requires homogeneity (that is, all the columns must be of same type.), time indices implicitly inferred as attributes or meta-information, series of same time length, and explicit missingness. By contrast, panel data are organised in two-dimensional array of heterogeneous column types where multiple study subjects are stacked and repeated for its time indices in a single column, due to commonly arisen unbalanced panels. This specification requires explicitly declared panel variable and index, which has been implemented in Stata's **tsset** and R package **plm**. This data organisation appears more advantageous than matrix in supporting explicit time index, multiple subjects of different time lengths.

This paper proposes a tidy and unified data structure and a modern data pipeline for storing, managing and analysing time series data, using a collection of fluent and fluid tools to help with exploitation in temporal context.

@@

@wickham2014tidy developed and formulated the conceptual framework of tidy data: (1) each variable forms a column; (2) each observation forms a row; (3) each type of observational unit forms a table. These principles attempt to standardise the mapping from the semantics of a dataset to its structure and facilitate data analysis in a coherent way. We shall extend "tidy data" into the time series domain.

@@

In SQL database, a primary key [@codd_relational_1970], which uniquely defines each record in a database table, is equivalent to the composition of "index" and "key".

@@

In experimental designs, a variable is crossed with another when every category of one variable co-occurs with every category of the other, while a variable is nested within another when each category of the former variable co-occurs with only one category of the latter. Regarding faceted plots in statistical graphics, @wilkinson2006grammar discussed the difference between nesting and crossing structure in depth. The distinction between two graphical layouts may appear subtle but convey completely different meanings of the data. The crossing example could be gender with marital status: married women, married men, single women, and single men. It takes all the possible combinations into account, resulting in 2 by 2 drawing panels in a graphical device. If a panel goes empty, there exists no observation for that category in the dataset, but it would be filled with observations. However, considering pregnancy status and gender, there are only three possible combinations: pregnant women, non-pregnant women, and non-pregnant men, since pregnant status is nested under the gender variable. In this case, any physical layout of such nesting structure must not occur to an impossible category.

It appears more useful in making this distinction in statistical analysis including visualisation and modeling, compared to data manipulation.

>>>>

---
title: "Tidy data structure to support exploration and modeling of temporal-context data"
author:
  - familyname: Wang
    othernames: Earo
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : earo.wang@monash.edu
    correspondingauthor: true
  - familyname: Cook
    othernames: Dianne
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : dicook@monash.edu
  - familyname: Hyndman
    othernames: Rob J
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : rob.hyndman@monash.edu
abstract: 'Temporal-context data is often rich with information and time formats, for example multiple observational units, different time lengths, heterogeneous data types, nested and crossed factors and etc. This work presents a cohesive and conceptual framework for organizing and manipulating temporal data, which in turn flows into visualization and modelling routines. Tidy data principles are applied and extended to temporal data: (1) mapping the semantics of a dataset into its physical layout, (2) an explicitly declared index variable representing time, (3) a "key" comprised of single or multiple variables to uniquely identify units over time, using a syntatical and user-oriented approach in which it imposes nested or crossed structures on the data. This tidy data representation most naturally supports thinking of operations on the data as building blocks, forming part of a "data pipeline" in time-based context. A sound pipeline practice facilitates a transparent and human readable workflow for analyzing temporal data. Applications are included to illustrate tidy temporal data structure, data pipeline ideas and usage. The architecture of tidy temporal data has been implemented in the R package **tsibble**.'
keywords: "temporal context, time series, data structure, R"
wpnumber: no/yr
jelcodes: C10,C14,C22
blind: false
cover: true
toc: false
bibliography: references.bib
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: no
    number_sections: yes
    citation_package: biblatex
---

@@

In the literature, time series and panel (longitudinal) data are common terms referred to temporal-context data, depending on the research fields. Researchers who are concerned with modelling large $T$ and small $N$ would name as "time series" (serial correlation); those who are interested in modelling small $T$ and large $N$ as "panel data" (asymptotic). The data format is two-dimensional array, but different modelling focuses lend the data input to different representations. A matrix is used to represent multivariate time series where each row represents observations measured at a time point and each column represents a series ("wide" form). This matrix representation requires homogeneity (that is, all the columns must be of same type.), time indices implicitly inferred as attributes or meta-information, series of same time length, and explicit missingness. By contrast, panel data are organised in rectangular form of heterogeneous column types where multiple study subjects are stacked and repeated for its time indices in a single column ("long" form), due to commonly arisen unbalanced panels. This specification requires explicitly declared panel variable and index, which has been implemented in Stata's time series module and R package **plm**. This data organisation appears more flexible than matrix in supporting explicit time index, multiple subjects of different time lengths, and implicit missing values.

Temporal data can often be aggregated in a manner that exhibits a nested or crossed structure, also known as hierarchical or grouped time series [@fpp]. For example, in a manufacturing setting, a company can add up every store’s sales by region, by state and by country, which gives a strictly hierarchical time series; alternatively they can group the sales for each product together based on common attributes such as store, brand, price range and so forth, which leads to a non-hierarchical structure---a grouped time series. The R package **hts** is the implementation of this type of time series. But it is frustrating to create and work with such data objects due to a machine-oriented interface.

@wickham2014tidy formulated the conceptual framework of tidy data: (1) each variable forms a column; (2) each observation forms a row; (3) each type of observational unit forms a table. These principles attempt to standardize the mapping from the semantics of a dataset to its structure and facilitate data analysis in a coherent way. A grammar of data manipulation.

This paper proposes a unified data representation of temporal-context data, by leveraging the "tidy data" principles. Built on a two-dimensional array of long format, observations and variables find and bridge their meanings in both physical and underlying representations.

This paper proposes a unified data representation and a modern data pipeline for storing, managing and analysing time series data, using a collection of fluent and fluid tools to help with exploitation in temporal context.

@@

@@

In SQL database, a primary key [@codd_relational_1970], which uniquely defines each record in a database, is equivalent to the composition of "index" and "key".

@@

In experimental designs, a variable is crossed with another when every category of one variable co-occurs with every category of the other, while a variable is nested within another when each category of the former variable co-occurs with only one category of the latter. It appears more useful in making this distinction in statistical analysis including visualisation and modeling, compared to data manipulation.
